---
title: "cran-comments"
output: github_document
---

## Reason for submission

`brokenstick 0.78.0` is a new package

## Test environments

* local OS X install, 10.15.7, R 4.0.2
* win-builder, using `devtools::check_win_devel()`
* rhub

### Local build

Package built by

```{r eval = FALSE}
library("devtools")
build(manual = TRUE)
```

```{bash eval=FALSE}
R CMD CHECK brokenstick_0.78.0.tar.gz

* checking re-building of vignette outputs ... WARNING
Error(s) in re-building vignettes:
--- re-building ‘brokenstick-article.Rmd’ using rmarkdown
Warning in file(con, "r") :
  cannot open file './_jss_txt.Rmd': No such file or directory
Quitting from lines 46-46 (./_jss_txt.Rmd) 
```

### WIN_DEVEL

`devtools::check_win_devel()` resulted in:

```
Status: OK
```

### RHUB

```{r eval=FALSE}
check_rhub()
```

The result is:

```
Build ID:	brokenstick_0.78.0.tar.gz-61572f5f668243edae78a65a4cb5259d
Platform:	Windows Server 2008 R2 SP1, R-devel, 32/64 bit
Submitted:	7 minutes 11 seconds ago
Build time:	6 minutes 32.2 seconds

...
> # case 3: only group
> predict(fit, test, group = c(11045, 11120, 999))
Error in loadNamespace(name) : there is no package called 'utf8'
Calls: <Anonymous> ... loadNamespace -> withRestarts -> withOneRestart -> doWithOneRestart
Execution halted

```

The solution of this problem is outside my reach. 

```
Build ID:	brokenstick_0.78.0.tar.gz-498bf3a054274ed69819b3c32a4d65af
Platform:	Ubuntu Linux 16.04 LTS, R-release, GCC
Submitted:	26 minutes 35.1 seconds ago
Build time:	26 minutes 14.8 seconds
```

Status: Success. Just a "new submission" note.

```
Build ID:	brokenstick_0.78.0.tar.gz-ce06a5c3e53d45f0bdf64595537d0c12
Platform:	Fedora Linux, R-devel, clang, gfortran
Submitted:	32 minutes 31.7 seconds ago
Build time:	31 minutes 52.9 seconds
```

Status: Success. Just a "new submission" note.


## Downstream dependencies

There are no downstream dependencies for this package.
