---
title: "cran-comments"
output: github_document
---

## Reason for submission

`brokenstick 0.78.0` is a new package

## Test environments

* local OS X install, 10.15.7, R 4.0.2
* win-builder, using `devtools::check_win_devel()`
* rhub

### Local build

Package built by

```{r eval = FALSE}
library("devtools")
build(manual = TRUE)
```

```{bash eval=FALSE}
R CMD CHECK brokenstick_0.78.0.tar.gz

Status: OK
```

### WIN_DEVEL

`devtools::check_win_devel()` resulted in:

```
Status: OK
```

### RHUB

```{r eval=FALSE}
check_rhub()
```

The result is:

```
── dscore 1.3.0: ERROR

  Build ID:   dscore_1.3.0.tar.gz-6c32ff0b664f45338fe9902216505e74
  Platform:   Windows Server 2008 R2 SP1, R-devel, 32/64 bit
  Submitted:  6m 29.3s ago
  Build time: 6m 26s

checking examples ...
  ** running examples for arch 'i386' ... ERROR
  Running examples in 'dscore-Ex.R' failed
  
  The error most likely occurred in:
  > ### Name: get_age_equivalent
  > base::assign(".ptime", proc.time(), pos = "CheckExEnv")
  > ### Title: Get age equivalents of items that have a difficulty estimate
  > 
  > ### ** Examples
  > ### Aliases: get_age_equivalent
  > 
  Error in loadNamespace(name) : there is no package called 'utf8'
```

The solution of this problem is outside my reach. 

```
── dscore 1.3.0: OK

Build ID:	dscore_1.3.0.tar.gz-56b28c78bf1045d386e3fda917897383
Platform:	Ubuntu Linux 16.04 LTS, R-release, GCC
Submitted:	28 minutes 49.8 seconds ago
Build time:	28 minutes 48.9 seconds
```

OK

```
── dscore 1.3.0: OK

Build ID:	dscore_1.3.0.tar.gz-b47f1ead4d784a42b7c747eae6e66f7c
Platform:	Fedora Linux, R-devel, clang, gfortran
Submitted:	34 minutes 23.5 seconds ago
Build time:	34 minutes 22.8 seconds
```

OK

```
── dscore 1.3.0: PREPERROR
Build ID:	dscore_1.3.0.tar.gz-4557c74d8e6c4c36a2635e6d04f75f8b
Platform:	Debian Linux, R-devel, GCC ASAN/UBSAN
Submitted:	59 minutes 50.4 seconds ago
Build time:	59 minutes 48.4 seconds

...
Build step 'Send files or execute commands over SSH' changed build result to SUCCESS
Pinging https://builder.r-hub.io/build/SUCCESS/dscore_1.3.0.tar.gz-4557c74d8e6c4c36a2635e6d04f75f8b/2020-05-12T14:53:59Z
{"status":"ok"}
Finished: SUCCESS
```

The banner says `PREPERROR` but the logs say `SUCCESS`

## Downstream dependencies

There are no downstream dependencies for this package.
