---
title: "cran-comments"
output: github_document
---

## Resubmission 4 to CRAN

* Wrap example section of `predict()` and `brokenstick()` in `\donttest{}` to reduce CPU time

## Resubmission 3 to CRAN

* We are now at version `brokenstick 2.1.0`
* Incorporates changes asked for by J Stat Soft (see package NEWS)
* Reduces package size from 3.7 Mb to 1.1 Mb by transferring large vignette to GitHub site

## Resubmission 2 to CRAN

* The submission changes a URL classified as 502 "Bad Gateway"
* The submission wraps examples in `\dontrun` from `predict()` to reduce CPU

## Resubmission 1 to CRAN

* This submission reduces package size from 9.2Mb to 3.7Mb by taking out temporary files that should not be in the distribution

## Test environments

* local OS X install, 12.2.1, R 4.1.3
* win-builder, using `devtools::check_win_devel()`
* rhub

### Local build

Package built by

```{r eval = FALSE}
library("devtools")
build(manual = TRUE)
```

```{bash eval=FALSE}
R CMD CHECK brokenstick_2.1.0.tar.gz

Status: OK
```

### WIN_DEVEL

`devtools::check_win_devel()` resulted in:

```
Note: CPU time > 10s for predict example section
```

### RHUB

```{r eval=FALSE}
rhub::check_for_cran()
```

`rhub::check_for_cran()` produces "SUCCESS" (three times) and one note:

```
*  checking examples ... NOTE
   Examples with CPU (user + system) or elapsed time > 5s
               user system elapsed
   predict     7.44   0.02    7.48
   brokenstick 4.96   0.02    5.15
```

Comment: 
Slightly above the threshold of 5 seconds. Both example sections demonstrate many capabilities of these functions to the end user. For clarity, I think these examples should remain as they are.


## Downstream dependencies

There are no downstream dependencies for this package.
