---
title: "Main functions"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Main functions}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 3.5)
```

## Objective

The vignette demonstrates the main functions in the `brokenstick` package: `brokenstick()`, `predict()` and `export`.

## Preliminaries

This vignette uses functions from the `rbokeh`, `hbgd` and `brokenstick` packages. 

```{r}
if (!require(rbokeh)) install.packages("rbokeh")
if (!require(hbgd)) devtools::install_github("HBGDki/hbgd")
require("brokenstick")
```


## Data set

The `smocc_hgtwgt` data in the `brokenstick` package contain the heights and weights of 1933 Dutch children measured on 10 visits at ages 0-2 years. Here we take the first 2000 records from the data.

```{r}
smc <- brokenstick::smocc_hgtwgt[1:2000, ]
head(smc, 3)
```

Another way to obtain these data is to use the `get_smocc_data()` function from the `github::HBGDki/hbgd` package, followed by a little post-processing:

```{r}
smc2 <- hbgd::get_smocc_data()[1:2000, ]
smc2$subjid <- as.numeric(as.character(smc2$subjid))
smc2$src <- as.character(smc2$src)
smc2$agedays <- round(smc2$agedays)
smc2$age <- round(smc2$agedays / 365.25, 4)
smc2$gagebrth <- smc2$ga * 7 + 3
smc2$etn <- as.character(smc2$etn)
smc2$birthwt <- smc2$bw
smc2$haz <- round(who_htcm2zscore(smc2$agedays, smc2$htcm, smc2$sex), 3)
smc2$waz <- round(who_wtkg2zscore(smc2$agedays, smc2$wtkg, smc2$sex), 3)
keep <- c("src", "subjid", "rec", "nrec",
          "age", "agedays", "sex", "etn",
          "gagebrth", "birthwt",
          "htcm", "haz", "wtkg", "waz")
smc2 <- smc2[, keep]
identical(smc, smc2)
```

Variable names and types conform to the HBGD definitions as implemented in `view_variables()`. 

## Exploratory analysis

Interactive plots of the height trajectories of three children can be made using the `rbokeh` package as follows:

```{r fig1}
ids <- c(10001, 10005, 10022)
d <- subset(smc, subjid %in% ids)
idx <- split(d, d$subjid)
figs <- lapply(idx, function(x) {
  figure(xlab = "Age (years)", ylab = "Length (cm)") %>%
  ly_who(x = seq(0, 750, by = 30), y_var = "htcm",
    x_trans = days2years, sex = x$sex[1], color = "green",
    p = 100 * pnorm(-c(2.5,2,1,0))) %>%
  ly_lines(days2years(x$agedays), x$htcm,
    col = "grey", hover = c(x$age, x$htcm)) %>%
  ly_points(days2years(x$agedays), x$htcm,
    col = "blue", hover = c(x$age, x$htcm), size = 6)
})
grid_plot(figs, same_axes = TRUE, simplify_axes = TRUE, width = 680, height = 300)
```

---

The plots gain more resolution when the Standard Deviation Scores (SDS) (or $Z$-values) are plotted. The column `haz` contains the $Z$-scores of height relative to the WHO standard.

```{r fig2}
figs <- lapply(idx, function(x) {
  figure(xlab = "Age (years)", ylab = "Length (SDS)") %>%
  ly_zband(x = days2years(seq(0, 750, by = 30)), z = -c(2.5,2,1,0)) %>%
  ly_lines(x$age, x$haz, col = "grey", hover = c(x$age, x$haz)) %>%
  ly_points(x$age, x$haz,
    col = "blue", hover = c(x$age, x$haz), size = 6)
})
grid_plot(figs, same_axes = TRUE, simplify_axes = TRUE, width = 680, height = 300)
```

---

## Broken stick model with two lines

The *broken stick model* describes a trajectory by a series of connected straight lines. We first calculate a rough approximation in the above trajectories using just two lines. The first line should start at birth and end at the age of exactly 1 years. The second line should span the period between 1 to 2 years. In addition, the lines must connect at the age of 1 year. We estimate this model as follows:

```{r fit1, cache = TRUE}
knots <- 0:2
fit1 <- brokenstick(y = smc$haz, 
					x = smc$age,
					subject = smc$subjid,
					knots = knots)
```

The `brokenstick()` function calls the `lmer()` function for actual model fitting. The `fit1` object has S4 class `brokenstick`, and collects the results from model fitting and some information related to the knot specification. For example, the `knots` argument to `brokenstick()` specifies knots at birth, 1 year and 2 years, respectively. In total, there are four knots:

```{r}
get_knots(fit1)
```

The last knot is equal to the maximum age, and is added automatically. The first and last knots may also be specified manually by the boundary argument. Proper knot placement is vital to obtaining a high quality model. The [vignette on knots placement](knotplacement.html) discusses the issue in more detail.

```{r}
class(fit1)
fit1
```

The fixed effects correspond to the mean of sample of children in the data. At birth, the children are on average 0.11 SD taller than the WHO standard. At the ages of 1 and 2 years, these estimates rise to 0.35 SD and 0.51 SD, respectively. The fact that these are positive and rising is not surprising. These are Dutch children, and the Dutch are known to be among the tallest in the world. Thus, at the age of 2 years, the Dutch are already about 0.5 SD taller than the WHO standard. The estimate for `x4` (0.87), corresponding to the maximum age in the data, is just there for technical reasons. It is based on extreme extrapolation beyond the data, and should be disregarded.

The column labeled `Std.Dev.` contains the standard deviations of the random effects. These numbers are expected to be around 1, since - under the assumption that the WHO standard appropriately describes this sample - the dependent variable follow a standard normal distribution with zero mean and standard deviation equal to one. The `Corr` component of the random effects lists the correlations between the broken stick estimates between the time point. Thus, the correlation between the estimates at birth (`x1`) and at 2 years (`x3`) is 0.51. The standard deviation of the residuals (within-person error) for this model is of 0.50. 



## Predicted values

The broken stick model defines predicted values as a compromise between the person's data values and the global mean. This compromise is called the *conditional mean* of the posterior distribution, it is calculated as the sum of fixed and random effects. The `predict()` function obtains predictions from the broken stick model. In order to obtain the predicted values for each measured data point for all children in the fitted model use

```{r}
p1 <- predict(fit1)
dim(p1)
head(p1, 4)
```

The observed $Z$-score is stored in column called `y`, and the corresponding broken stick estimates are stored in the column called `yhat`. The column `knot` is set to `FALSE` if `y` is actually observed. We obtain the connected lines according to the broken stick model by obtaining predictions at the break ages by specifying the `at = "knots"` argument.

```{r}
p2 <- predict(fit1, at = "knots")
head(p2, 4)
```

Break ages are distinguihsed from observed ages by the `knot` column, which is now `TRUE`. We may obtain both types of estimates simultaneously for all children by 

```{r}
pr <- predict(fit1, x = get_knots(fit1))
head(pr, 15)
```

We plot the original data and the broken stick estimates jointly by `xyplot()` from the `lattice` package by

```{r fig3}
d <- subset(pr, subjid %in% ids & x <= 2.2)
idx <- split(d, d$subjid)
figs <- lapply(idx, function(x) {
  figure(xlab = "Age (years)", ylab = "Length (SDS)") %>%
  ly_zband( x = days2years(seq(0, 750, by = 30)), z = -c(2.5,2,1,0)) %>%
  ly_lines( x = x$x[!x$knot], y = x$y[!x$knot], color = "grey") %>%
  ly_points(x = x$x[!x$knot], x$y[!x$knot], color = "blue", size = 6) %>%
  ly_lines( x = x$x[x$knot], y = x$yhat[x$knot], col = "darkred") %>%
  ly_points(x = x$x[x$knot], x$yhat[x$knot], col = "red", size = 6) 
  })
grid_plot(figs, same_axes = TRUE, simplify_axes = TRUE, 
          width = 680, height = 300)
```

---

## Extension to nine lines

We now refine the model in the first two years by adding a knot for each age at which a visit was scheduled. This model can be run as

```{r fit2, cache = TRUE}
# 10 scheduled visits
knots <- round(c(0, 1, 2, 3, 6, 9, 12, 15, 18, 24)/12, 4)
boundary <- c(0, 3)
fit2 <- brokenstick(y = smc$haz, 
					x = smc$age,
					subject = smc$subjid,
					knots = knots,
					boundary = boundary)
```

The model contains 11 random effects. This optimization problem is more difficult. The optimization routine issues a number of warnings related to the number of random effects relative to the number of observations, and the complexity of the optimization problem. These warnings may indicate degeneracies in the numerical solution, and are generally be less frequent for large samples. 

As before, predicted values can be obtained by

```{r}
pr <- predict(fit2, x = get_knots(fit_hgt))
head(pr, 4)
```

```{r echo=FALSE}
d <- subset(pr, subjid %in% ids & x <= 2.2)
idx <- split(d, d$subjid)
figs <- lapply(idx, function(x) {
  figure(xlab = "Age (years)", ylab = "Length (SDS)") %>%
  ly_zband( x = days2years(seq(0, 750, by = 30)), z = -c(2.5,2,1,0)) %>%
  ly_lines( x = x$x[!x$knot], y = x$y[!x$knot], color = "grey") %>%
  ly_points(x = x$x[!x$knot], x$y[!x$knot], color = "blue", size = 6) %>%
  ly_lines( x = x$x[x$knot], y = x$yhat[x$knot], col = "darkred") %>%
  ly_points(x = x$x[x$knot], x$yhat[x$knot], col = "red", size = 6) 
  })
grid_plot(figs, same_axes = TRUE, simplify_axes = TRUE, 
          width = 680, height = 300)
```

---

The figure plots the nine line model, with break points at the scheduled ages (indicated by the grey dashed lines). The model now fits the data very well. Formal model assessment can be done with standard `lmer` diagnostic plots. See the vignette on *Model diagnostics* for more detail [NOTE: vignette still has to be written]. 

The broken stick model is sensitive to the location of the knots. Some advice about knot placement: Evidently, using more knots can improve the fit to the data. However, adding an excessive number of knots will substantially increase calculation time, and create solutions that are less stable. Try using not more than the (average) number of data points per person. Also, knot placement requires some care. Here, the knots are added at the scheduled visits, which is a reasonable strategy for obtaining good predictions as there are generally many observations in the neighourbood of the scheduled ages. In contrast, placing knots at ages where the data are very sparse (e.g. exactly in-between visits) may result in bad predictions and should be avoided. 

## Explained variance

The proportion of the variance of the outcome explained by the two-line model is

```{r}
var(fitted(fit1), na.rm = TRUE) / var(smc$haz, na.rm = TRUE)
```

For the second model we get

```{r}
var(fitted(fit2), na.rm = TRUE) / var(smc$haz, na.rm = TRUE)
```

so the nine-line broken stick model explains about `r round(100 * var(fitted(fit2), na.rm = TRUE) / var(smc$haz, na.rm = TRUE))` percent of the variance of the height SDS.


## Publishing your model

The `fit2` object can be stored by `save()` for future use in R. However, a more parsimonious and convenient way to reuse the model is to export its parameter estimates, and store only these. As a matter of fact, prediction using the broken stick model requires only the knot locations, the fixed effect estimates, the variance/covariance matrix of the random effects, and the size of the residual variance. The `export()` function will collect this information from the fitted model, and store it as a list.

```{r}
# export the broken stick models
exp <- export(fit2)
attributes(exp)
lapply(exp, round, 2)
```

Such estimates can be published as a longitudinal growth reference.

## Prediction using a published broken stick reference

Suppose we have four measured a new child and want to obtain predictions according to the published model stored in `export_hgt`. The following code calculates the predictions.

```{r}
# four height measurement on new child
x <- c(0, 0.12, 0.32, 0.62, 1.1)
y <- c(-1.2, -1.8, -1.7, -1.9, -2.1)

# prediction at ages x
atx <- predict(exp, y = y, x = x)
atx
# prediction at the knots
atknots <- predict(exp, y = y, x = x, at = "knots")
head(atknots)
```

The predicted values are found such that the published model fits best to the given age and height values. The figure below plots the observed data and the predicted data at the break ages for the new child.

```{r echo = FALSE, fig.align = "center"}
pr <- rbind(atx, atknots)
d <- subset(pr, x <= 2.2)
idx <- list(d)
figs <- lapply(idx, function(x) {
  if (is.null(x)) return(NULL)
  figure(xlab = "Age (years)", ylab = "Length (SDS)") %>%
  ly_zband( x = days2years(seq(0, 750, by = 30)), z = -c(2.5,2,1,0)) %>%
  ly_lines( x = x$x[!x$knot], y = x$y[!x$knot], color = "grey") %>%
  ly_points(x = x$x[!x$knot], x$y[!x$knot], color = "blue", size = 10) %>%
  ly_lines( x = x$x[x$knot], y = x$yhat[x$knot], col = "red") %>%
  ly_points(x = x$x[x$knot], x$yhat[x$knot], col = "red", size = 10) 
  })

grid_plot(figs)
```

