---
title: "Overview of main functions"
author: "Stef van Buuren"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette
bibliography: [bibliography.bib]
biblio-style: apalike
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Overview main functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.retina = 2, 
                      fig.width = 7, fig.height = 3.5)
```

## Objective

This vignette demonstrates three major functions in the `brokenstick` package: `brokenstick()`, `predict()` and `plot()`.

## Set up libraries

This vignette uses functions from the `rbokeh`, `hbgd` and `brokenstick` packages. 

```{r message=FALSE}
if (!require(rbokeh)) devtools::install_github("hafen/rbokeh")
if (!require(hbgd)) devtools::install_github("HBGDki/hbgd")
require("hbgd")
require("brokenstick")
```


## Load data set

The `smocc_50` data in the `brokenstick` package contain the heights of 50 Dutch children measured on 10 visits at ages 0-2 years.

```{r}
smc <- brokenstick::smocc_50
head(smc, 3)
```


## Plot trajectories

Let us first plot some of the individual growth curves. We may do so after fitting a trivial broken stick model, and plot the trajectories of the three selected children as follows: 

```{r measurements}
fit <- brokenstick(y = smc$hgt, x = smc$age, subjid = as.factor(smc$subjid))
ids <- c(10001, 10005, 10022)
plot(fit, ids = ids, zband = FALSE, height = 350, width = 225, ylab = "Length (cm)", pkg = "ggplot")
```

The plots gain more resolution when the Standard Deviation Scores (SDS) (or $Z$-values) are plotted. The column `hgt.z` contains the $Z$-scores of height relative to the WHO standard.

```{r zscore}
fit0 <- brokenstick(y = smc$hgt.z, x = smc$age, subjid = as.factor(smc$subjid))
plot(fit0, ids = ids, pkg = "ggplot")
```

The same type of plot can be made by `bokeh`:

## Fit broken stick model with two lines

The *broken stick model* describes a trajectory by a series of connected straight lines. We first calculate a rough approximation in the above trajectories using just two lines. The first line starts at birth and end at the age of exactly 1 years. The second line spans the period between 1 to 2 years. In addition, the lines must connect at the age of 1 year. We estimate this model as follows:

```{r fit2, cache = TRUE}
fit2 <- brokenstick(y = smc$hgt.z, x = smc$age, subjid = smc$subjid, knots = 1:2)
```

The fitted model can now be plotted by 

```{r plotfit2}
plot(fit2, ids = ids, x_trim = c(0, 2.2))
```


The fixed effects correspond to the mean of sample of children in the data, whereas the correlation between ages is modelled by the random effects. The interpretation of the estimates is discussed in the [vignette for prediction](prediction.html).

## Extend to nine lines

The plot shows that the two-line model is fairly crude. We now refine the model in the first two years by adding a knot for each age at which a visit was scheduled. This model can be run as

```{r fit9, cache = TRUE, warning=FALSE}
# 10 scheduled visits
knots <- round(c(0, 1, 2, 3, 6, 9, 12, 15, 18, 24)/12, 4)
fit9 <- brokenstick(y = smc$hgt.z, x = smc$age, subjid = smc$subjid, knots = knots)
```

The model contains 11 random effects. This optimization problem is more difficult, so it takes longer to run. In addition, it is common that the optimization routines issue a number of warnings related to the number of random effects relative to the number of observations. In general, these warnings do not seem to affect the quality of the solution.

As before, the observed and fitted trajectories can be plotted by

```{r plotfit9}
plot(fit9, ids = ids, x_trim = c(0, 2.2))
```

The broken stick model now fits the observed data quite close.

## Obtain predicted values

The `predict()` function allows us to obtain various types of predictions from the broken stick model. The simplest call

```{r}
p1 <- predict(fit2)
head(p1, 4)
```

produces the predicted value (in `yhat`) for each measured data point for all children used to fit the model in the `long` output format. The observed $Z$-score is stored in column called `y`, and the corresponding broken stick estimates are stored in the column called `yhat`. The column `knot` is set to `FALSE` if `y` is actually observed. 

We can obtain the locations at which the lines connect (i.e. at the knots) by specifying the `at = "knots"` argument, e.g.

```{r}
p2 <- predict(fit2, at = "knots")
head(p2, 4)
```

Note that the column `knot` is now `TRUE`. Note there are also knots at ages 0.00 and 2.68 years. These are boundary knots, and added by the `brokenstick()` function. By default, the boundary knots span the age range in the data. The estimate for the knot at the maximum age has no useful interpretation. We may obtain estimates at both knots and observed ages by setting `at = "both"`. 

## Explained variance

The proportion of the variance of the outcome explained by the two-line model is

```{r}
get_pev(fit2)
```

For the second model we get

```{r}
get_pev(fit9)
```

so the nine-line broken stick model explains about `r round(100 * get_pev(fit9))` percent of the variance of the height SDS.


## Subject level analysis

Suppose we are interest in knowing the effect of sex, gestational age and birth weight on the height SDS at the age of 2 years. This is an analysis at the subject level. The `get_subject_data()` function from the `hbgd` package extracts the variables that vary only over subjects.

```{r eval=FALSE}
subj <- get_subject_data(smc)
head(subj, 3)
```

The broken stick estimates per subject from the nine line solution are obtained by

```{r eval=FALSE}
bs <- predict(fit9, at = "knots", output = "broad")
dim(bs)
head(round(bs, 2), 3)
```

The names of the columns correspond to the ages at which the knots were placed. The effect of the subject's sex, gestational age and birth weight on the height SDS at the age of 2 years (here denoted by the variable named `2`) can be estimated as 

```{r lm1, eval = FALSE}
data <- cbind(subj, bs)
fit1_lm <- lm(`2` ~ sex + ga + I(bw / 1000), data = data)
summary(fit1_lm)
```

Note that the analysis shows there is a substantial effect of birth weight. Of course, it might be that birth weight is directly related to height at the age of 2 years. Alternatively, the relation could be mediated by birth length. The following model adds birth length (the variable named `0`) to the model: 

```{r lm2, eval=FALSE}
fit2_lm <- lm(`2` ~ sex + ga + I(bw / 1000) + `0`, data = data)
summary(fit2_lm)
```

The effect of birth length on length at age 2 is very strong. There is no separate effect of birth weight anymore, so this analysis suggests that the relation between birth weight and length at age 2 can be explained by their mutual associations to birth length.

## Conclusion

This vignette illustrated the use of the `brokenstick()`, `plot()` and `predict()` functions. Other vignettes highlight various other capabilities of the package. 

* [Fit and predict](prediction.html)
    + Data exploration
    + Conversion back and forth to the Z-score scale
    + Export and publish a broken stick model
    + Predict growth curve of new subjects
    + Assess the quality of the model
 
 
* [Model formulation](model.html)
    + Definition of the model
    + Interpretation of the model


* [Perfect model](perfectmodel.html)
    + Properties of the perfect model
    + Estimating time-to-time correlations

* [Knot placement](knotplacement.html)
    + Knot placement advice
