---
title: "Help for old friends"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Help for old friends}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(brokenstick)
```

# Background

Versions after `brokenstick 0.62.1` (May 2020) have a substantially changed interface. In particular, later versions (starting with `brokenstick 0.70.0`) include the following changes: 

1. `brokenstick` adopts the `tidymodels` philosophy. It is now possible to fit a model using five different interfaces. There is no need anymore the hardcode variable names in the source data.

2. Version `0.70.0` introduces a new estimation method, the Kasim-Raudenbush sampler. The new method is more flexible and faster than `lme4::lmer()` when the number of knots is large.

3. Version `0.70.0` introduces two simple correlation models that may be used to smooth out the variance-covariance matrix of the random effects.

4. The definition of the `brokenstick` class has changed. Objects of class `brokenstick` do no longer store the training data. 

5. The `brokenstick_export` class is retired.

6. The `predict()` function is fully rewritten as has now a new interface. Since the `brokenstick` class does not store the training data anymore, the `predict()` function now obtains a `new_data` argument. Syntax that worked for `brokenstick` package before `0.70.0` does not work anymore and should be updated. The `shape` argument replaces the `output` argument.

7. The `plot()` function is rewritten, and now requires a `new_data` specification.

8. Replaced functions: `brokenstick()` replaces `fit_brokenstick()`, `predict.brokenstick()` replaces `predict.brokenstick_export()`, `get_r2()` replaces `get_pev()`

9. Removed functions: `get_data()`, `get_X()`, `export()`

# Install legacy version

We recommend changing your code to reflect the above changes. If needed, the old version `0.62.1` can be installed as

```{r eval=FALSE}
library(devtools)
install_github("stefvanbuuren/brokenstick@V0.62.1")
```

# Examples

## Example 1: Fit model

Fit model, old version:

```{r fit_old, eval=FALSE}
data <- brokenstick::smocc_200
fit <- brokenstick(y = data$hgt.z, x = data$age, subjid = data$subjid)
```

Fit model, version `0.70.0` and higher:

```{r fit_new}
data <- brokenstick::smocc_200

# formula interface
fit1 <- brokenstick(hgt.z ~ age | id, data)

# XY interface - numeric vector
fit2 <- with(data, brokenstick(age, hgt.z, id))

# Recipes data.frame interface
library(recipes)
rec <- recipe(data,
              vars = c("hgt.z", "age", "id"),
              roles = c("outcome", "predictor", "group"))
fit3 <- brokenstick(rec, data)

# XY interface - data.frame
fit4 <- with(data, brokenstick(data.frame(age), hgt.z, id))

# XY interface - matrix
tt <- as.matrix(data[, c(1, 2, 7)])
fit5 <- brokenstick(tt[, "age", drop = FALSE],
                    tt[, "hgt.z", drop = FALSE],
                    tt[, "id", drop = FALSE])
```

All fitted models are identical.

## Example 2: Predict model

Predict model, old version:

```{r predict_old, eval = FALSE}
# predict at observed data
p1 <- predict(fit)

# predict at knots
p2 <- predict(fit, at = "knots")

# predict at both observed data and knots
p3 <- predict(fit, at = "both")

# predict knots, broad version
p4 <- predict(fit, at = "knots", output = "broad")
```


Predict model, new version:

```{r predict_new}
# predict at observed data
p1 <- predict(fit1, data)

# predict at knots
p2 <- predict(fit1, data, x = "knots")

# predict at both observed data and knots
p3 <- predict(fit1, data, x = "knots", strip_data = FALSE)

# predict knots, broad matrix
p4 <- predict(fit1, data, x = "knots", shape = "wide")
```


## Example 3: Plot model

Plot trajectories, old version 

```{r plot_old, eval = FALSE}
ids <- c(10001, 10005, 10022)
plot(fit, ids = ids)
```

Plot trajectories, new version 

```{r plot_new, fig.height=3, fig.width=7}
ids <- c(10001, 10005, 10022)
plot(fit1, data, group = ids)
```

## Example 4: Explained variance

Proportion explained variance, old version

```{r explain_old, eval=FALSE}
get_pev(fit)
```


```{r explain_new}
get_r2(fit1, data)
```

